\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2021-11-29 main.Rmd
tweak section 2 more + some cleaning up

\newpage
## Create a cubble in the nested form

@@

In the `cubble` header, you can read the name of the `key` variable, bbox, and also the name of variable nested in the `ts` column. In this example, the spatial identifier is `id` and the number in the bracket means there are 5 unique `id` in this dataset. The bbox in the second row gives the range of the coordinates. The temporal variables are all nested in the `ts` column, but it could be useful to know the name these variables. The third row in the cubble header shows these names and in this example this includes: precipitation, `prcp`, maximum temperature, `tmax`, and minimum temperature, `tmin`.

@@

From a created cubble in the nested form, analysts may want to directly manipulate the temporal variables. This would require switching to the long form using `stretch()`. The verb `stretch()` switch the cubble from the nested form into a long form. Under the hood, it first extracts the spatial variables into a separate tibble to store in the attribute `spatial` and then unnests the `ts` column to show the temporal content:

@@

Notice here that the third line in the header is changed to reflect the spatial variables stored. This is a format suitable for computing time-wise variables.

@@

Manipulation on the spatial and temporal dimension can be an iterative process. Many times, we may decide to go back to the nested form after some temporal manipulation. The verb to switch a long cubble back to the nested form is `tamp()`:

@@

As an output to be supplied to further visualisation or modelling, analysts would usually like the spatial and temporal variables to be in the same table. `migrate()` moves the spatial variables in the attribute `spatial` into the long form cubble.

@@

## Support on hierarchical structure

@@

## Support on interactive graphics

## Integrating into a tidy workflow

\begin{center}
\begin{tabular}{ | m{5em} | m{15cm}| } 
\textbf{mutate} \\
\textbf{filter}\\
\textbf{summarise} \\
\textbf{select} \\
\textbf{arrange} \\
\textbf{rename} \\
\textbf{left\_join} \\
\textbf{group\_by} \\
\textbf{ungroup}\\
slice family & \textbf{slice\_head}, \textbf{slice\_tail}, \textbf{slice\_sample}, \textbf{slice\_min} and \textbf{slice\_max} \\
\end{tabular}
\end{center}

@@

```{r}
gly <- glyphs(
  tmax_long2,
  x_major = "longitude",
  y_major = "latitude",
  x_minor = "month",
  y_minor = "tmax",
  height = 0.6,
  width = 1
)

@@

```{r}
cluster_long <- cluster_nested %>%
  stretch(ts) %>%
  group_by(wk) %>%
  summarise(prcp = sum(prcp, na.rm = TRUE)) %>%
  migrate(cent_long, cent_lat)

@@

```{r fig.height = 5, fig.cap = "shtishisa"}
station_long <- station_nested %>%
  stretch(ts) %>%
  migrate(cluster)
station_long %>%
  ggplot(aes(x = wk, y = prcp, group = id)) +
  geom_line(alpha = .3) +
  facet_wrap(vars(cluster), scales = "free_y", ncol = 4) +
  theme_bw()
```

@@

```{r}
# this part still needs some fixing
tas_latlong <- station_nested %>%
  filter(lat < -40) %>%
  mutate(p = max(ts$prcp)) %>%
  strip_rowwise() %>%
  filter(p == max(p))

@@

```{r}
river <- weatherdata::water %>%
  stretch() %>%
  select(date, Water_course_level) %>%
  tamp() %>%
  mutate(type = "river")

@@

The output from temporal matching is also a cubble, with additional column `.dist` and `.group` inherent from spatial matching and `n_match` for the number of matched temporal peaks. Then you can use this output to plot the location of match or to look at the series: 
```{r out.width="100%", fig.cap="..."}
p1 <- plot_map(vic_map) +
  geom_point(
    data = res,
    aes(x = long, y = lat, color = type)
  ) +
  ggrepel::geom_label_repel(
    data = res %>% filter(type == "river"),
    aes(x = long, y = lat, label = .group)
  ) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("Victoria") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

@@

```{r eval = FALSE}
set.seed(123)
tmax_data <- weatherdata::climate_full %>%
  slice_sample(n = 50) %>%
  stretch() %>%
  mutate(ym = tsibble::yearmonth(date)) %>%
  group_by(ym) %>%
  summarise(
    tmax = mean(tmax, na.rm = TRUE),
    tmin = mean(tmin, na.rm = TRUE),
    prcp = sum(prcp, na.rm = TRUE)
  ) %>%
  ungroup(ym) %>%
  mutate(
    year = as.factor(year(ym)),
    month = as.factor(month(ym)),
    prcp = sqrt(prcp + 0.001)
  ) %>%
  tamp() %>%
  mutate(median_tmax = median(ts$tmax, na.rm = TRUE)) %>%
  stretch() %>%
  migrate(median_tmax, long, lat, name) %>%
  highlight_key(~id)

p1 <- tmax_data %>%
  ggplot(aes(
    x = month, y = tmax, group = interaction(year, id),
    color = median_tmax, label = name
  )) +
  geom_line() +
  facet_wrap(vars(fct_reorder(name, -median_tmax))) +
  theme(legend.position = "none")

@@

Many spatial and spatio-temporal data structures have been developed by the R-spatial team for both raster and vector spatial data. For vector spatial data, which is the focus of this paper, `sf` [@pebesma2018simple] represents spatial vector information with simple features: points, lines, polygons and their multiples. Various `st_` function are designed to manipulate these features based on their geometric relationships. For spatio-temporal data,  `stars` [@stars] can represent both raster and vector data using multi-dimensional array. However, the underlying array structure can be difficult to operate for data analysts who are more familiar with a flat 2D data frame structure used by the tidyverse ecosystem.

@@

# A new data structure for spatio-temporal data

**The main difficulty and challenge**  
The main difficulty in visualising this type of data is to show information in both space and time dimension with the proper level of details without information overflow. This would sometimes require aggregating the time dimension into the proper level or slicing the data into a reasonable number of subset for display. In this sense, a data structure that regulates the manipulation spatio-temporal data will benefit the analysis workflow. While many implementations focus on manipulating and visualising pure spatial or temporal data, there are not sufficient tools to deal with spatio-temporal data. The purpose of this paper is to introduce a spatio-temporal vector data structure for data analysis in R.

To work with spatio-temporal data, analysts can choose to either work separately on each dimension or join the two sets together, however, each approach has its own problem: While is is natural to work separately on each sheet (since spatial and temporal operations usually don't overlap), analysts will need to manually keep the other data frame up to date. For example, the following pseudo code illustrates the scenario where once the spatial dataset is filtered for those within Victoria, the temporal dataset needs to be manually updated to reflect this spatial filter.

```{r eval = FALSE}
spatial_new <- spatial %>% filter(SITES_IN_VICTORIA)
temporal_new <- temporal %>% filter(id %in% spatial_new$id)
```

If analysts choose to join the spatial and temporal data together, the joined dataset could be too large since each spatial variable will be repeated at each time stamp for each site. Also, recordings of the site ID from different data sources can be slightly different from each other, causing a painful checking and cleaning of site IDs before the join.

@@

### Basics

- `stretch`: nest to long form
- `tamp`: long to nest form
- `migrate`: move selected spatial variables to the long form.
- `add_dscrb_prct`: summary stats for missingness

dplyr compatibility:

  - mutate, filter, summarise, select, arrange
  - group and ungroup: group_by, ungroup
  - slice family

### Combine two cubbles

  - match river and weather gauges data
  - involve combining two cubbles
  - join operations combine the two together by appending more rows but what we really want is to bind rows.
  - bind rows also doesn't work since we want to bind only when there' s a matching????
  - introduce bind_join

### Hierarchical structure in cubble

  - hierarchical is common.
  - Given examples.
  - Essence: switch between different levels
  - introduce `switch_key`

Daily climate data (prcp, tmax, and tmin) from RNOAA - lots of stations across Australia

An exploratory data analysis questions: What's the climate profile look like in Australia

  - General features: Any general trend/ fluctuation in prcp, tmax, and tmin?
  - Local features: Any station stands out from the crowd?

>>>>

## Create a cubble in the nested form

@@

There are a few information in the `cubble` header: the name of the `key` variable, bbox, and also the name of variable nested in the `ts` column. In this example, each site is identifier is `id` and the number in the bracket means there are 5 unique `id` in this dataset. The bbox in the second row gives the range of the coordinates. The temporal variables are all nested in the `ts` column, but it could be useful to know the name these variables. The third row in the cubble header shows these names and in this example this includes: precipitation, `prcp`, maximum temperature, `tmax`, and minimum temperature, `tmin`.

@@

The long cubble is suitable to manipulate the time dimension of the data. The function `stretch()` switches the nested cubble into the long cubble by first extracts all the spatial variables into a separate tibble and store in the `spatial` attribute and then unnests the `ts` column:

@@

Notice here that the third line in the header now shows the name of  spatial variables rather than the temporal variables.

@@

Manipulation on the spatial and temporal dimension can be an iterative process. Many times, analysts will need to go back and forth between the nested and long cubble. The `stretch()` function introduced in the previous section switches a nested cubble into a long cubble and function `tamp()` is its inverse function to switch a long cubble back to the nested cubble:

@@

As a final data output for modelling or visualisation, spatio-temporal data is usually expected to be in a single table. Function `migrate()` moves the spatial variables from the `spatial` attribute into the long cubble:

@@

In this workflow described above, data objects come into cubble in the nested form, then various operations on the spatial and temporal dimension can go back and forth between the nested and long form, and finally, the data will come out of cubble in the long form for further modelling or visualisation.

\begin{center}
\begin{tabular}{ | m{5em} | m{15cm}| } 
basics & \textbf{mutate}, \textbf{filter}, \textbf{summarise}, \textbf{select}, \textbf{arrange}, \textbf{rename}, \textbf{left\_join} \\
grouping &  \textbf{group\_by}, \textbf{ungroup}\\
slice family & \textbf{slice\_head}, \textbf{slice\_tail}, \textbf{slice\_sample}, \textbf{slice\_min} and \textbf{slice\_max} \\
\end{tabular}
\end{center}

# Advanced features/ considerations

## Hierarchical structure

@@

## Interactive graphics

## Data fusion and matching

@@

```{r echo = FALSE}
gly <- glyphs(
  tmax_long2,
  x_major = "longitude",
  y_major = "latitude",
  x_minor = "month",
  y_minor = "tmax",
  height = 0.6,
  width = 1
)

@@

```{r echo = FALSE}
cluster_long <- cluster_nested %>%
  stretch(ts) %>%
  group_by(wk) %>%
  summarise(prcp = sum(prcp, na.rm = TRUE)) %>%
  migrate(cent_long, cent_lat)

@@

```{r fig.height = 5, fig.cap = "shtishisa", echo = FALSE}
station_long <- station_nested %>%
  stretch(ts) %>%
  migrate(cluster)
station_long %>%
  ggplot(aes(x = wk, y = prcp, group = id)) +
  geom_line(alpha = .3) +
  facet_wrap(vars(cluster), scales = "free_y", ncol = 4) +
  theme_bw()
```

@@

```{r echo = FALSE}
# this part still needs some fixing
tas_latlong <- station_nested %>%
  filter(lat < -40) %>%
  mutate(p = max(ts$prcp)) %>%
  strip_rowwise() %>%
  filter(p == max(p))

@@

```{r echo = FALSE}
river <- weatherdata::water %>%
  stretch() %>%
  select(date, Water_course_level) %>%
  tamp() %>%
  mutate(type = "river")

@@

The output from temporal matching is also a cubble, with additional column `.dist` and `.group` inherent from spatial matching and `n_match` for the number of matched temporal peaks. Then you can use this output to plot the location of match or to look at the series: 
```{r echo = FALSE, out.width="100%", fig.cap="..."}
p1 <- plot_map(vic_map) +
  geom_point(
    data = res,
    aes(x = long, y = lat, color = type)
  ) +
  ggrepel::geom_label_repel(
    data = res %>% filter(type == "river"),
    aes(x = long, y = lat, label = .group)
  ) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("Victoria") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

@@

```{r eval = FALSE}
set.seed(123)
dt <- weatherdata::climate_full %>%
  slice_sample(n = 50) %>% 
  mutate(name = str_replace(name, "\\s*\\([^\\)]+\\)", ""))

tmax_data <- dt %>%
  stretch() %>%
  mutate(ym = tsibble::yearmonth(date)) %>%
  group_by(ym) %>%
  summarise(
    tmax = mean(tmax, na.rm = TRUE),
    tmin = mean(tmin, na.rm = TRUE),
    prcp = sum(prcp, na.rm = TRUE)
  ) %>%
  ungroup(ym) %>%
  mutate(
    year = as.factor(year(ym)),
    month = as.factor(month(ym)),
    prcp = sqrt(prcp + 0.001)
  ) %>%
  tamp() %>%
  mutate(median_tmax = median(ts$tmax, na.rm = TRUE)) %>%
  stretch() %>%
  migrate(median_tmax, long, lat, name) %>%
  highlight_key(~id)
```

```{r eval = FALSE, echo = FALSE}
p1 <- tmax_data %>%
  ggplot(aes(
    x = month, y = tmax, group = interaction(year, id),
    color = median_tmax, label = name
  )) +
  geom_line() +
  scale_x_discrete(breaks = seq(1, 12, 3)) + 
  facet_wrap(vars(fct_reorder(name, -median_tmax))) +
  theme(legend.position = "none")

@@

@@

@@

### 2021-11-29 main.tex
tweak section 2 more + some cleaning up

In the \texttt{cubble} header, you can read the name of the \texttt{key}
variable, bbox, and also the name of variable nested in the \texttt{ts}
column. In this example, the spatial identifier is \texttt{id} and the
number in the bracket means there are 5 unique \texttt{id} in this
dataset. The bbox in the second row gives the range of the coordinates.
The temporal variables are all nested in the \texttt{ts} column, but it
could be useful to know the name these variables. The third row in the
cubble header shows these names and in this example this includes:
precipitation, \texttt{prcp}, maximum temperature, \texttt{tmax}, and
minimum temperature, \texttt{tmin}.

@@

From a created cubble in the nested form, analysts may want to directly
manipulate the temporal variables. This would require switching to the
long form using \texttt{stretch()}. The verb \texttt{stretch()} switch
the cubble from the nested form into a long form. Under the hood, it
first extracts the spatial variables into a separate tibble to store in
the attribute \texttt{spatial} and then unnests the \texttt{ts} column
to show the temporal content:

@@

Manipulation on the spatial and temporal dimension can be an iterative
process. Many times, we may decide to go back to the nested form after
some temporal manipulation. The verb to switch a long cubble back to the
nested form is \texttt{tamp()}:

@@

As an output to be supplied to further visualisation or modelling,
analysts would usually like the spatial and temporal variables to be in
the same table. \texttt{migrate()} moves the spatial variables in the
attribute \texttt{spatial} into the long form cubble.

@@

\hypertarget{support-on-hierarchical-structure}{%
\subsection{Support on hierarchical
structure}\label{support-on-hierarchical-structure}}

@@

\hypertarget{support-on-interactive-graphics}{%
\subsection{Support on interactive
graphics}\label{support-on-interactive-graphics}}

\hypertarget{integrating-into-a-tidy-workflow}{%
\subsection{Integrating into a tidy
workflow}\label{integrating-into-a-tidy-workflow}}

\begin{center}
\begin{tabular}{ | m{5em} | m{15cm}| } 
\textbf{mutate} \\
\textbf{filter}\\
\textbf{summarise} \\
\textbf{select} \\
\textbf{arrange} \\
\textbf{rename} \\
\textbf{left\_join} \\
\textbf{group\_by} \\
\textbf{ungroup}\\
slice family & \textbf{slice\_head}, \textbf{slice\_tail}, \textbf{slice\_sample}, \textbf{slice\_min} and \textbf{slice\_max} \\
\end{tabular}
\end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> gly <- glyphs(
+   tmax_long2,
+   x_major = "longitude",
+   y_major = "latitude",
+   x_minor = "month",
+   y_minor = "tmax",
+   height = 0.6,
+   width = 1
+ )
R> 
R> nsw_vic <- ozmaps::abs_ste %>%
+   filter(NAME %in% c("Victoria", "New South Wales"))
R> 
R> plot_map(nsw_vic, fill = "transparent") +
+   geom_path(data = gly, aes(gx, gy,
+     group = interaction(id, group),
+     color = group
+   )) +
+   scale_color_brewer(palette = "Dark2") + 
+   theme_bw() + 
+   coord_sf(xlim = c(140, 155)) + 
+   theme(legend.position = "bottom")
\end{CodeInput}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cluster_long <- cluster_nested %>%
+   stretch(ts) %>%
+   group_by(wk) %>%
+   summarise(prcp = sum(prcp, na.rm = TRUE)) %>%
+   migrate(cent_long, cent_lat)
R> 
R> gly <- GGally::glyphs(cluster_long,
+   x_major = "cent_long", x_minor = "wk",
+   y_major = "cent_lat", y_minor = "prcp",
+   height = 2, width = 4
+ )
R> 
R> state_map <- rmapshaper::ms_simplify(ozmaps::abs_ste, keep = 2e-3)
R> plot_map(state_map) +
+   geom_text(
+     data = cluster_nested,
+     aes(x = cent_long, y = cent_lat, label = cluster)
+   ) +
+   geom_path(
+     data = gly,
+     aes(x = gx, y = gy, group = gid)
+   )
\end{CodeInput}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> station_long <- station_nested %>%
+   stretch(ts) %>%
+   migrate(cluster)
R> station_long %>%
+   ggplot(aes(x = wk, y = prcp, group = id)) +
+   geom_line(alpha = .3) +
+   facet_wrap(vars(cluster), scales = "free_y", ncol = 4) +
+   theme_bw()
\end{CodeInput}
\begin{figure}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # this part still needs some fixing
R> tas_latlong <- station_nested %>%
+   filter(lat < -40) %>%
+   mutate(p = max(ts$prcp)) %>%
+   strip_rowwise() %>%
+   filter(p == max(p))
R> 
R> state_map <- rmapshaper::ms_simplify(ozmaps::abs_ste, keep = 2e-3)
R> plot_map(state_map) +
+   geom_point(data = station_nested, aes(x = long, y = lat), size = 0.5) +
+   geom_sf(data = cluster_nested, aes(geometry = ll_cluster), fill = "transparent") +
+   geom_point(data = tas_latlong, aes(x = long, y = lat), col = "red")
\end{CodeInput}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> river <- weatherdata::water %>%
+   stretch() %>%
+   select(date, Water_course_level) %>%
+   tamp() %>%
+   mutate(type = "river")
R> 
R> climate <- weatherdata::climate_full %>%
+   filter(between(stringr::str_sub(id, 7, 8), 76, 90)) %>%
+   stretch() %>%
+   filter(lubridate::year(date) == 2020) %>%
+   tamp() %>%
+   mutate(type = "climate")
R> 
R> vic_map <- rmapshaper::ms_simplify(ozmaps::abs_ste %>% filter(NAME == "Victoria"))
R> plot_map(vic_map) +
+   geom_point(
+     data = dplyr::bind_rows(river, climate),
+     aes(x = long, y = lat, color = type)
+   ) +
+   scale_color_brewer(palette = "Dark2") +
+   theme_bw()
\end{CodeInput}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> p1 <- plot_map(vic_map) +
+   geom_point(
+     data = res,
+     aes(x = long, y = lat, color = type)
+   ) +
+   ggrepel::geom_label_repel(
+     data = res %>% filter(type == "river"),
+     aes(x = long, y = lat, label = .group)
+   ) +
+   scale_color_brewer(palette = "Dark2") +
+   ggtitle("Victoria") +
+   theme_minimal() +
+   theme(
+     panel.grid.major = element_blank(),
+     panel.grid.minor = element_blank(),
+     legend.position = "bottom"
+   )
R> 
R> res_long <- res %>%
+   stretch(ts) %>%
+   migrate(.group, type) %>%
+   mutate(prcp = scale(prcp)[, 1]) %>%
+   switch_key(.group) %>%
+   migrate(type)
R> 
R> p2 <- res_long %>%
+   ggplot(aes(
+     x = date, y = prcp,
+     group = id, color = type
+   )) +
+   geom_line() +
+   facet_wrap(vars(.group)) +
+   scale_color_brewer(palette = "Dark2", guide = "none") +
+   theme_bw() +
+   labs(x = "week") +
+   scale_x_date(date_labels = "%b")
R> 
R> p1 | p2
\end{CodeInput}
\begin{figure}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(123)
R> tmax_data <- weatherdata::climate_full %>%
+   slice_sample(n = 50) %>%
+   stretch() %>%
+   mutate(ym = tsibble::yearmonth(date)) %>%
+   group_by(ym) %>%
+   summarise(
+     tmax = mean(tmax, na.rm = TRUE),
+     tmin = mean(tmin, na.rm = TRUE),
+     prcp = sum(prcp, na.rm = TRUE)
+   ) %>%
+   ungroup(ym) %>%
+   mutate(
+     year = as.factor(year(ym)),
+     month = as.factor(month(ym)),
+     prcp = sqrt(prcp + 0.001)
+   ) %>%
+   tamp() %>%
+   mutate(median_tmax = median(ts$tmax, na.rm = TRUE)) %>%
+   stretch() %>%
+   migrate(median_tmax, long, lat, name) %>%
+   highlight_key(~id)
R> 
R> p1 <- tmax_data %>%
+   ggplot(aes(
+     x = month, y = tmax, group = interaction(year, id),
+     color = median_tmax, label = name
+   )) +
+   geom_line() +
+   facet_wrap(vars(fct_reorder(name, -median_tmax))) +
+   theme(legend.position = "none")
R> 
R> state_map <- rmapshaper::ms_simplify(ozmaps::abs_ste, keep = 2e-3)
R> p2 <- tmax_data %>%
+   ggplot() +
+   geom_sf(data = state_map, aes(geometry = geometry)) +
+   geom_point(aes(x = long, y = lat, color = median_tmax, label = name)) +
+   theme_void()
R> 
R> bscols(
+   ggplotly(p1, width = 800, height = 800, tooltip = "label") %>%
+     highlight(on = "plotly_selected", off = "plotly_deselect", color = "red"),
+   ggplotly(p2, width = 800, height = 800, tooltip = "label") %>%
+     highlight(on = "plotly_selected", off = "plotly_deselect", color = "red")
+ )
\end{CodeInput}
\end{CodeChunk}

@@

Many spatial and spatio-temporal data structures have been developed by
the R-spatial team for both raster and vector spatial data. For vector
spatial data, which is the focus of this paper, \texttt{sf}
\citep{pebesma2018simple} represents spatial vector information with
simple features: points, lines, polygons and their multiples. Various
\texttt{st\_} function are designed to manipulate these features based
on their geometric relationships. For spatio-temporal data,
\texttt{stars} \citep{stars} can represent both raster and vector data
using multi-dimensional array. However, the underlying array structure
can be difficult to operate for data analysts who are more familiar with
a flat 2D data frame structure used by the tidyverse ecosystem.

@@

\hypertarget{a-new-data-structure-for-spatio-temporal-data}{%
\section{A new data structure for spatio-temporal
data}\label{a-new-data-structure-for-spatio-temporal-data}}

\textbf{The main difficulty and challenge}\\
The main difficulty in visualising this type of data is to show
information in both space and time dimension with the proper level of
details without information overflow. This would sometimes require
aggregating the time dimension into the proper level or slicing the data
into a reasonable number of subset for display. In this sense, a data
structure that regulates the manipulation spatio-temporal data will
benefit the analysis workflow. While many implementations focus on
manipulating and visualising pure spatial or temporal data, there are
not sufficient tools to deal with spatio-temporal data. The purpose of
this paper is to introduce a spatio-temporal vector data structure for
data analysis in R.

To work with spatio-temporal data, analysts can choose to either work
separately on each dimension or join the two sets together, however,
each approach has its own problem: While is is natural to work
separately on each sheet (since spatial and temporal operations usually
don't overlap), analysts will need to manually keep the other data frame
up to date. For example, the following pseudo code illustrates the
scenario where once the spatial dataset is filtered for those within
Victoria, the temporal dataset needs to be manually updated to reflect
this spatial filter.

\begin{CodeChunk}
\begin{CodeInput}
R> spatial_new <- spatial %>% filter(SITES_IN_VICTORIA)
R> temporal_new <- temporal %>% filter(id %in% spatial_new$id)
\end{CodeInput}
\end{CodeChunk}

If analysts choose to join the spatial and temporal data together, the
joined dataset could be too large since each spatial variable will be
repeated at each time stamp for each site. Also, recordings of the site
ID from different data sources can be slightly different from each
other, causing a painful checking and cleaning of site IDs before the
join.

@@

\hypertarget{basics}{%
\subsubsection{Basics}\label{basics}}

\begin{itemize}
\tightlist
\item
  \texttt{stretch}: nest to long form
\item
  \texttt{tamp}: long to nest form
\item
  \texttt{migrate}: move selected spatial variables to the long form.
\item
  \texttt{add\_dscrb\_prct}: summary stats for missingness
\end{itemize}

dplyr compatibility:

\begin{itemize}
\tightlist
\item
  mutate, filter, summarise, select, arrange
\item
  group and ungroup: group\_by, ungroup
\item
  slice family
\end{itemize}

\hypertarget{combine-two-cubbles}{%
\subsubsection{Combine two cubbles}\label{combine-two-cubbles}}

\begin{itemize}
\tightlist
\item
  match river and weather gauges data
\item
  involve combining two cubbles
\item
  join operations combine the two together by appending more rows but
  what we really want is to bind rows.
\item
  bind rows also doesn't work since we want to bind only when there' s a
  matching????
\item
  introduce bind\_join
\end{itemize}

\hypertarget{hierarchical-structure-in-cubble}{%
\subsubsection{Hierarchical structure in
cubble}\label{hierarchical-structure-in-cubble}}

\begin{itemize}
\tightlist
\item
  hierarchical is common.
\item
  Given examples.
\item
  Essence: switch between different levels
\item
  introduce \texttt{switch\_key}
\end{itemize}

\hypertarget{examples-1}{%
\section{Examples}\label{examples-1}}

Daily climate data (prcp, tmax, and tmin) from RNOAA - lots of stations
across Australia

An exploratory data analysis questions: What's the climate profile look
like in Australia

\begin{itemize}
\tightlist
\item
  General features: Any general trend/ fluctuation in prcp, tmax, and
  tmin?
\item
  Local features: Any station stands out from the crowd?
\end{itemize}

>>>>

There are a few information in the \texttt{cubble} header: the name of
the \texttt{key} variable, bbox, and also the name of variable nested in
the \texttt{ts} column. In this example, each site is identifier is
\texttt{id} and the number in the bracket means there are 5 unique
\texttt{id} in this dataset. The bbox in the second row gives the range
of the coordinates. The temporal variables are all nested in the
\texttt{ts} column, but it could be useful to know the name these
variables. The third row in the cubble header shows these names and in
this example this includes: precipitation, \texttt{prcp}, maximum
temperature, \texttt{tmax}, and minimum temperature, \texttt{tmin}.

@@

The long cubble is suitable to manipulate the time dimension of the
data. The function \texttt{stretch()} switches the nested cubble into
the long cubble by first extracts all the spatial variables into a
separate tibble and store in the \texttt{spatial} attribute and then
unnests the \texttt{ts} column:

@@

Manipulation on the spatial and temporal dimension can be an iterative
process. Many times, analysts will need to go back and forth between the
nested and long cubble. The \texttt{stretch()} function introduced in
the previous section switches a nested cubble into a long cubble and
function \texttt{tamp()} is its inverse function to switch a long cubble
back to the nested cubble:

@@

As a final data output for modelling or visualisation, spatio-temporal
data is usually expected to be in a single table. Function
\texttt{migrate()} moves the spatial variables from the \texttt{spatial}
attribute into the long cubble:

@@

In this workflow described above, data objects come into cubble in the
nested form, then various operations on the spatial and temporal
dimension can go back and forth between the nested and long form, and
finally, the data will come out of cubble in the long form for further
modelling or visualisation.

\begin{center}
\begin{tabular}{ | m{5em} | m{15cm}| } 
basics & \textbf{mutate}, \textbf{filter}, \textbf{summarise}, \textbf{select}, \textbf{arrange}, \textbf{rename}, \textbf{left\_join} \\
grouping &  \textbf{group\_by}, \textbf{ungroup}\\
slice family & \textbf{slice\_head}, \textbf{slice\_tail}, \textbf{slice\_sample}, \textbf{slice\_min} and \textbf{slice\_max} \\
\end{tabular}
\end{center}

\hypertarget{advanced-features-considerations}{%
\section{Advanced features/
considerations}\label{advanced-features-considerations}}

\hypertarget{hierarchical-structure}{%
\subsection{Hierarchical structure}\label{hierarchical-structure}}

@@

\hypertarget{interactive-graphics}{%
\subsection{Interactive graphics}\label{interactive-graphics}}

\hypertarget{data-fusion-and-matching}{%
\subsection{Data fusion and matching}\label{data-fusion-and-matching}}

@@

\begin{CodeChunk}

@@

\begin{CodeChunk}

@@

@@

\begin{CodeChunk}

@@

\begin{CodeChunk}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(123)
R> dt <- weatherdata::climate_full %>%
+   slice_sample(n = 50) %>% 
+   mutate(name = str_replace(name, "\\s*\\([^\\)]+\\)", ""))
R>  
R> tmax_data <- dt %>%
+   stretch() %>%
+   mutate(ym = tsibble::yearmonth(date)) %>%
+   group_by(ym) %>%
+   summarise(
+     tmax = mean(tmax, na.rm = TRUE),
+     tmin = mean(tmin, na.rm = TRUE),
+     prcp = sum(prcp, na.rm = TRUE)
+   ) %>%
+   ungroup(ym) %>%
+   mutate(
+     year = as.factor(year(ym)),
+     month = as.factor(month(ym)),
+     prcp = sqrt(prcp + 0.001)
+   ) %>%
+   tamp() %>%
+   mutate(median_tmax = median(ts$tmax, na.rm = TRUE)) %>%
+   stretch() %>%
+   migrate(median_tmax, long, lat, name) %>%
+   highlight_key(~id)
\end{CodeInput}
\end{CodeChunk}

@@

@@

@@
